---
title: "XRD Crystallinity -- Summary Report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`"
output:
  html_document:
    toc: false
    theme: default
params:
  result: NULL
---

```{r tcr-report-setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 7.5, fig.height = 4.5)
library(textileCrystR)
result <- params$result
report_kind <- if (inherits(result, "textile_cryst_batch")) {
  "segal_batch"
} else if (inherits(result, "textile_cryst_peakarea")) {
  "peakarea"
} else {
  "segal_single"
}
```

This report was generated by `textileCrystR`. Every method it implements is
an **empirical, relative** X-ray diffraction index, not an absolute
crystallinity -- see the *Limitations* section below, and note that
different methods are never averaged together (see `?textileCrystR`).

```{r tcr-report-segal-single-summary, eval = report_kind == "segal_single", results = "asis"}
p <- result$points
cat(sprintf("**Method:** Segal peak height\n\n"))
cat(sprintf("**Sample:** %s\n\n", if (is.na(result$sample)) "(unnamed)" else result$sample))
cat(sprintf("**Segal Crystallinity Index:** %.2f %%\n\n", result$ci))
knitr::kable(
  data.frame(Point = p$point, `Position requested (deg)` = round(p$position_requested, 3),
             `Position used (deg)` = round(p$position_used, 3),
             Intensity = round(p$intensity, 1), check.names = FALSE),
  caption = "Positions and intensities used."
)
```

```{r tcr-report-segal-single-plot, eval = report_kind == "segal_single"}
plot(result)
```

```{r tcr-report-segal-single-warnings, eval = report_kind == "segal_single" && length(result$warnings) > 0, results = "asis"}
cat("**Warnings:**\n\n")
for (w in result$warnings) cat(sprintf("- %s\n", w))
```

```{r tcr-report-segal-batch-summary, eval = report_kind == "segal_batch", results = "asis"}
tab <- as.data.frame(result)
cat(sprintf("**Method:** Segal peak height\n\n"))
cat(sprintf("**%d sample(s) analyzed.**\n\n", nrow(tab)))
knitr::kable(
  data.frame(Sample = tab$sample, `Segal CI (%)` = round(tab$ci, 2),
             `I200 position` = round(tab$i200_position, 2),
             `Iam position` = round(tab$iam_position, 2),
             Warnings = tab$n_warnings, check.names = FALSE),
  caption = "Segal Crystallinity Index by sample."
)
```

```{r tcr-report-segal-batch-plot, eval = report_kind == "segal_batch"}
plot(result)
```

```{r tcr-report-peakarea-summary, eval = report_kind == "peakarea", results = "asis"}
cat(sprintf("**Method:** Peak area / deconvolution (%s profiles)\n\n", result$model))
cat(sprintf("**Sample:** %s\n\n", if (is.na(result$sample)) "(unnamed)" else result$sample))
cat(sprintf("**Crystallinity Index:** %.2f %%\n\n", result$ci))
cat(sprintf("**Fit quality:** R-squared = %.4f, RMSE = %.2f\n\n", result$r_squared, result$rmse))
tab <- rbind(result$peaks[, c("label", "center", "height", "fwhm", "area", "pct_of_crystalline")],
            result$amorphous[, c("label", "center", "height", "fwhm", "area", "pct_of_crystalline")])
knitr::kable(
  data.frame(Peak = tab$label, `Center (deg)` = round(tab$center, 2), Height = round(tab$height, 1),
             FWHM = round(tab$fwhm, 3), Area = round(tab$area, 1),
             `% of crystalline` = ifelse(is.na(tab$pct_of_crystalline), "--",
                                         sprintf("%.1f%%", tab$pct_of_crystalline)),
             check.names = FALSE),
  caption = "Fitted peaks and amorphous hump."
)
if (result$background$model != "none") {
  cat(sprintf("\n\nBackground: %s (intercept %.1f%s)\n\n", result$background$model,
             result$background$intercept,
             if (result$background$model == "linear") sprintf(", slope %.4f", result$background$slope)
             else ""))
}
```

```{r tcr-report-peakarea-plot, eval = report_kind == "peakarea", fig.height = 6}
plot_peak_fit(result)
```

```{r tcr-report-peakarea-warnings, eval = report_kind == "peakarea" && length(result$warnings) > 0, results = "asis"}
cat("**Warnings:**\n\n")
for (w in result$warnings) cat(sprintf("- %s\n", w))
```

## Method

```{r tcr-report-method-text, results = "asis"}
if (report_kind %in% c("segal_single", "segal_batch")) {
  cat("```\nCI = (I200 - Iam) / I200 * 100\n```\n\n")
  cat("Intensities are read at the reported positions, interpolating between ",
      "measured points where a position was not measured exactly. No background ",
      "correction is applied.\n", sep = "")
} else {
  cat("```\n", result$metadata$equation, "\n```\n\n", sep = "")
  cat("Crystalline peaks and the amorphous hump are fit **simultaneously** as one ",
      "multi-component nonlinear model (this is necessary since they overlap); ",
      "crystallinity is the crystalline peaks' combined area over the total area.\n",
      sep = "")
}
```

## Limitations

```{r tcr-report-limitations-text, results = "asis"}
if (report_kind %in% c("segal_single", "segal_batch")) {
  cat("The Segal index depends on peak overlap, crystallite size, cellulose ",
      "polymorph, preferred orientation, background handling, the exact positions ",
      "chosen for I200 and Iam, and sample preparation, and is not appropriate for ",
      "partially mercerized cotton when cellulose II positions are used.\n", sep = "")
} else {
  cat("Peak-area crystallinity depends on the chosen peak-shape model, the number ",
      "and starting positions of crystalline peaks, the amorphous window, and (if ",
      "fit) the background model -- it is not an absolute crystalline fraction. ",
      "Compare fits made the same way, and report the model and positions used ",
      "alongside the value.\n", sep = "")
}
```

See `?textileCrystR` for full details and literature references (Segal et al.,
1959; Nam et al., 2016; Salem et al., 2023).

*Generated by `textileCrystR` `r utils::packageVersion("textileCrystR")`.*
