---
title: "Getting Started with textileCrystR"
author: "textileCrystR"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with textileCrystR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2, dpi = 96)
library(textileCrystR)
```

This vignette is a practical, start-to-finish tour of `textileCrystR` as a
piece of software: install it, point it at a file, get a number and a plot,
and export the results -- without needing to know the internals of
`segal_ci()` or write any plotting code yourself. For the scientific detail
behind the Segal index itself (its equation, presets and limitations) see
`vignette("segal-cotton-validation")` and `?textileCrystR`; this vignette is
about *using* the package.

## 1. Install

```r
install.packages("textileCrystR")

# Optional, only needed for the interactive dashboard (Section 8):
install.packages(c("shiny", "rmarkdown"))
```

```{r library}
library(textileCrystR)
```

## 2. The one-line quick start

The fastest way to use the package is `analyze_textile_file()`: point it at
a file and it reads, validates, calculates and plots in one call. We'll use
the experimental cotton XRD data set shipped with the package.

```{r quickstart}
path <- system.file("extdata", "cotton.csv", package = "textileCrystR")
res <- analyze_textile_file(path, sample = "0%/1")
```

`res` is the same kind of result `segal_ci()` returns, so everything you can
do with it there works here too:

```{r quickstart-inspect}
res
summary(res)
```

## 3. Loading your own data

Real files are messier than a tidy example, so `import_xrd_file()` (used
internally by `analyze_textile_file()`, and also usable on its own) is
forgiving about format: it auto-detects the delimiter (comma, tab,
semicolon or plain whitespace), whether there's a header row, and -- from
header names like "Angle"/"2-theta" and "Counts"/"Intensity" -- which
column is which.

```{r import-plain}
tt <- seq(5, 40, by = 0.05)
y <- 300 + 4000 * exp(-((tt - 22.7) / 1.1)^2) + 1500 * exp(-((tt - 16) / 1.5)^2)
f <- tempfile(fileext = ".csv")
writeLines(c("Angle,Counts", paste(round(tt, 3), round(y, 1), sep = ",")), f)

xrd <- import_xrd_file(f)
head(xrd)
```

If your file has no header, or names its columns in a way the package
doesn't recognize, tell it explicitly which columns to use with `cols`:

```{r import-cols}
f2 <- tempfile(fileext = ".csv")
writeLines(paste(round(tt, 3), round(y, 1), sep = ","), f2)  # no header at all
xrd2 <- import_xrd_file(f2, cols = c(1, 2))
head(xrd2)
```

### When something's wrong with the file

Rather than a cryptic parsing error, `import_xrd_file()` explains what it
saw and what to try. For example, a file where the columns are ambiguous:

```{r import-error, error = TRUE}
bad <- tempfile(fileext = ".csv")
writeLines(c("Sample,Reading", "501,3200", "502,3305", "503,3100"), bad)
import_xrd_file(bad)
```

The message names the problem (the header didn't name a recognizable
2-theta column, and the values in column 1 don't look like one either) and
says exactly what to do (`cols = c(...)`). The same applies to
`analyze_textile_file()`, a nonexistent path, an unreadable format, and so
on -- see `?import_xrd_file` for the full list of checks.

## 4. Choosing I200 and Iam: presets, manual positions, and sensitivity

By default, `analyze_textile_file()` uses the `"cellulose_I_segal"` preset
(I200 = 22.7 degrees, Iam = 18.0 degrees). List the available presets, each
with its material, reference and limitations:

```{r presets}
segal_presets()[, c("preset", "material", "i200", "iam")]
```

Use a different preset, or set positions manually:

```{r positions}
res_preset <- analyze_textile_file(path, sample = "0%/1", preset = "cotton_validation",
                                   plot = FALSE)
res_manual <- analyze_textile_file(path, sample = "0%/1", i200 = 22.55, iam = 17.9,
                                   plot = FALSE)
c(preset = res_preset$ci, manual = res_manual$ci)
```

Positions matter more than they might seem. `segal_sensitivity()` shows how
much the index moves across a grid of nearby positions:

```{r sensitivity}
d <- import_xrd_file(path)
d0 <- d[d$sample == "0%/1", ]
sens <- segal_sensitivity(d0$two_theta, d0$intensity)
sens
```

Report the positions you used alongside any Segal value -- this is why every
`textile_cryst` result keeps them (`res$points`), and `summary()` prints them.

## 5. Working with multiple samples at once

Files like `cotton.csv` hold several samples. Leave `sample` unset to
analyze all of them and get a `textile_cryst_batch`:

```{r batch}
batch <- analyze_textile_file(path, preset = "cotton_validation", plot = FALSE)
batch
```

`as.data.frame()` gives the tidy table for further analysis (joining onto
your own metadata, plotting with your own tools, etc.):

```{r batch-df}
as.data.frame(batch)
```

`plot()` on a batch draws every sample's own zoomed-in window, side by side,
with its I200/Iam points marked -- handy for spotting a sample whose peak
has drifted from a fixed position, as `100%/1` has here:

```{r batch-plot}
plot(batch)
```

## 6. Exporting results

Both single-sample and batch calls can write their outputs directly, so a
script can go from raw file to finished deliverables in one call:

```{r export}
out_plot <- tempfile(fileext = ".png")
out_report <- tempfile(fileext = ".html")
analyze_textile_file(path, sample = "0%/1", preset = "cotton_validation",
                     plot = FALSE, plot_file = out_plot, report_file = out_report)
file.exists(out_plot)
file.exists(out_report)
```

`report_file` (needs the `rmarkdown` package) writes a self-contained HTML
report with the plot, the positions and intensities used, any warnings, and
the same limitations section as `?textileCrystR` -- suitable for sharing
with a collaborator who doesn't use R. The exact same template is used by
the interactive dashboard's "Download summary report" button (Section 8),
so the two are interchangeable.

## 7. Peak area / deconvolution: a second method

Everything so far has used the Segal method. `textileCrystR` also implements
a second, independent method -- peak area / deconvolution -- which jointly
fits several sharp crystalline peaks and one broad amorphous hump, and
reports crystallinity as the crystalline peaks' combined area over the total
area:

```{r peakarea-fit}
fit <- fit_crystalline_peaks(d0$two_theta, d0$intensity, sample = "0%/1")
fit
```

```{r peakarea-plot}
plot_peak_fit(fit)
```

By default this fits four literature-cited cellulose I crystalline peaks
((1-10), (110), (200), (004)) plus one amorphous hump, using pseudo-Voigt
profiles and a linear background term -- real XRD patterns carry a
background (air scatter, detector dark counts) that a bare "peaks + hump"
model has no way to represent on its own. The model, peak centers, amorphous
window and background handling can all be set explicitly; see
`?fit_crystalline_peaks`.

This is a genuinely independent estimate, not a refinement of the Segal
value -- the two are not expected to agree, and are never averaged together:

```{r peakarea-vs-segal}
segal_res <- segal_ci(d0$two_theta, d0$intensity, preset = "cotton_validation")
c(segal = segal_res$ci, peak_area = fit$ci)
```

See `?textileCrystR` for why independent XRD methods disagree and how to
interpret that.

## 8. Going interactive: the dashboard

For point-and-click use across one or many files and samples, with either
method, launch the bundled Shiny app:

```r
run_textile_app()
```

This needs the `shiny` and `rmarkdown` packages (Section 1); the check for
them, and a clear message if either is missing, happens only when you call
`run_textile_app()`, not when you load the package. A few things worth
knowing before you launch it:

* **Upload one or more files** at once. Every sample across every file is
  pooled into a single selector, labelled by file name only when more than
  one file is uploaded (keeping the common single-file case uncluttered).
* **Subtract an estimated background before analysis**, optionally -- a
  non-parametric SNIP-based estimate (`estimate_xrd_background()`),
  independent of either method's own settings. Off by default; when on,
  single-sample mode shows a preview (raw pattern, estimated background,
  corrected pattern) so you can judge the window before trusting it.
* **Pick a method** -- Segal or peak area / deconvolution -- from the
  sidebar; the rest of the sidebar adapts to that method's own controls
  (positions and presets for Segal; model, peak centers, amorphous window
  and background for peak area).
* **Single-sample mode** is a live tuning surface: for Segal, drag the
  sliders or click directly on the peak/trough in the plot; for peak area,
  adjust settings and press "Run fit" -- fitting is a real optimization, not
  an instant calculation, so it isn't tied to every slider movement.
* **Batch mode** analyzes any number of the pooled samples (or all of them)
  together with the current method and settings, producing a results table,
  a comparison chart, and a downloadable CSV. A sample that fails (for
  example, its measured range doesn't cover the chosen positions) is
  reported individually without losing the ones that succeeded.
* The plot/report download buttons (single-sample mode) produce exactly the
  files `plot_file` and `report_file` do in Section 6, for whichever method
  is currently selected.

## Where to go next

* `?textileCrystR` -- the full list of scientific limitations and references.
* `?segal_ci`, `?segal_presets`, `?segal_sensitivity` -- the Segal building
  blocks, if you want more control than `analyze_textile_file()` gives you.
* `?fit_crystalline_peaks`, `?fit_amorphous_hump`, `?peak_fit_summary` -- the
  peak-area / deconvolution building blocks (Section 7).
* `vignette("segal-cotton-validation")` -- the Segal method itself, validated
  end to end against the shipped experimental cotton data set.
