---
title: "Segal Crystallinity Analysis of Experimental Cotton XRD Data"
author: "textileCrystR"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Segal Crystallinity Analysis of Experimental Cotton XRD Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2,
                      dpi = 96)
library(textileCrystR)
library(ggplot2)
```

This vignette walks through a complete Segal analysis of a real experimental
data set: seven X-ray diffraction patterns of cotton containing different
gravimetric ratios of mercerized and raw cotton. `textileCrystR` also
implements a second, independent method (peak area / deconvolution -- see
`vignette("getting-started")`); this vignette focuses specifically on the
Segal peak-height crystallinity index

$$\mathrm{CI} = \frac{I_{200} - I_{am}}{I_{200}} \times 100 .$$

Throughout, the result is called the **Segal Crystallinity Index**. It is an
empirical, relative XRD index, not an absolute crystallinity (Segal et al.,
1959; Nam et al., 2016; Salem et al., 2023). The composition labels of the
series (0 %/1 ... 100 %/1) are metadata of the samples; they are never used in
any calculation and are not a ground-truth crystallinity scale.

## 1. Loading the data

`cotton.csv` stores each pattern in two adjacent columns (2θ, intensity) under
a sample name. `read_xrd_pairs()` reads the file exactly as stored.

```{r load}
path <- system.file("extdata", "cotton.csv", package = "textileCrystR")
xrd <- read_xrd_pairs(path)
head(xrd, 3)
table(xrd$sample)[c("0%/1", "20%/1", "40%/1", "50%/1", "60%/1", "80%/1", "100%/1")]
```

## 2. Validation

`validate_crystallinity_input()` checks every pattern before use: numeric
types, equal lengths, missing values, the 0-180° range, ordering and duplicated
angles. It never changes measured intensities. Here all seven patterns pass
without any action being needed.

```{r validate}
samples <- c("0%/1", "20%/1", "40%/1", "50%/1", "60%/1", "80%/1", "100%/1")
checks <- do.call(rbind, lapply(samples, function(s) {
  d <- xrd[xrd$sample == s, ]
  v <- validate_crystallinity_input(d$two_theta, d$intensity)
  data.frame(sample = s, points = v$n_used, from = v$range[1], to = v$range[2],
             step = round(v$median_step, 3), sorted_by_package = v$was_sorted,
             actions = length(v$messages))
}))
knitr::kable(checks, caption = "Input checks for the seven cotton patterns.")
```

## 3. Segal calculation

We use the validation convention I200 = 22.6° and Iam = 18.0° 2θ. The pattern
was measured every 0.01°, so both positions coincide with measured points and
the measured intensities are used exactly. If a requested position were not
measured, the intensity would be interpolated (see `?segal_points`).

```{r single}
d <- xrd[xrd$sample == "0%/1", ]
res <- segal_ci(d$two_theta, d$intensity, i200 = 22.6, iam = 18.0, sample = "0%/1")
res
```

## 4. Inspecting the result

The result object keeps everything needed to reproduce the number: the
positions requested and used, the intensities, the neighbouring measured angles,
the method settings and the validated pattern.

```{r inspect}
summary(res)
res$points
as.data.frame(res)
segal_points(d$two_theta, d$intensity, i200 = 22.62, iam = 18.013)[
  c("i200_position", "i200_intensity", "iam_position", "iam_intensity", "interpolation")]
```

## 5. Plotting

`plot()` returns an ordinary `ggplot` object, so it can be modified freely.

```{r plot}
p <- plot(res)
p
```

```{r plot-modified}
p + theme_bw() + labs(title = "Raw cotton (0%/1)") +
  geom_hline(yintercept = res$points$intensity[2], linetype = "dotted")
```

At 22.6° the I200 point lies on the low-angle side of the (200) maximum, which
is close to 22.9° for this sample. This is why the choice of position matters
(Section 8).

## 6. All seven cotton samples

```{r all}
fits <- lapply(samples, function(s) {
  d <- xrd[xrd$sample == s, ]
  segal_ci(d$two_theta, d$intensity, i200 = 22.6, iam = 18.0, sample = s)
})
all_res <- do.call(rbind, lapply(fits, as.data.frame))
all_res$composition_label_pct <- as.numeric(sub("%.*$", "", all_res$sample))
knitr::kable(
  transform(all_res[, c("sample", "ci", "i200_intensity", "iam_intensity")],
            ci = round(ci, 2), i200_intensity = round(i200_intensity, 1),
            iam_intensity = round(iam_intensity, 1)),
  col.names = c("Sample", "Segal CI (%)", "I200 (22.6°)", "Iam (18.0°)"),
  caption = "Segal Crystallinity Index of the seven cotton patterns.")
```

```{r composition-plot}
ggplot(all_res, aes(composition_label_pct, ci)) +
  geom_line(colour = "#A9B4D0") + geom_point(colour = "#243070", size = 3) +
  scale_x_continuous(breaks = all_res$composition_label_pct) +
  labs(x = "Composition label of the sample series (%)", y = "Segal Crystallinity Index (%)",
       subtitle = "Composition is a sample label, not a crystallinity scale") +
  theme_minimal()
```

The index decreases from 71.19 % to 56.96 % across the series but not
monotonically (40 %/1 is higher than 20 %/1). The gravimetric ratio should not
be expected to give a linear crystallinity scale, and the series is compared
here only as an observed relationship.

## 7. Independent reference comparison

The same equation was evaluated independently in a spreadsheet, which reads the
measured rows of the raw data directly with `INDEX()` and has no interpolation or
R code in the loop (`data-raw/textileCrystR_cotton_validation.xlsx` in the source
package; the values it calculated are shipped as
`inst/extdata/reference_cotton_segal.csv`). This is an *implementation check* of
the same Segal equation on the same data, not a different scientific method.

```{r reference}
ref <- read.csv(system.file("extdata", "reference_cotton_segal.csv", package = "textileCrystR"))
cmp <- merge(all_res[, c("sample", "ci")], ref[, c("sample", "ci_reference")], by = "sample")
cmp <- cmp[match(samples, cmp$sample), ]
cmp$abs_difference <- abs(cmp$ci - cmp$ci_reference)
knitr::kable(data.frame(Sample = cmp$sample,
                        `CI from R (%)` = round(cmp$ci, 6),
                        `CI from reference (%)` = round(cmp$ci_reference, 6),
                        `Absolute difference (pp)` = format(cmp$abs_difference, digits = 2),
                        check.names = FALSE),
             caption = "textileCrystR against the spreadsheet calculation.")
```

The maximum absolute difference is `r format(max(cmp$abs_difference), digits = 2)`
percentage points (floating-point level). The workbook was prepared for this
project, so the check verifies the code path, not the choice of positions.

**Literature context (not a benchmark to match).** Nam et al. (2016) report
Segal CIs of 84.7 % (control cotton) and 68.8 % (mercerized cotton) for their
own samples and simulations, and Salem et al. (2023, Table 2) list 85-87 % for
cotton linter by the Segal peak-height method. Values from different instruments,
sample forms, background treatments and position choices are not comparable, and
the values here are not expected to match them. What can be said is that the
direction of change, a lower Segal CI for the sample at the mercerized end of the series, is
in line with the drop reported by Nam et al. (2016).

## 8. Sensitivity analysis

`segal_sensitivity()` evaluates the index for I200 = 22.2, 22.4, 22.6, 22.8,
23.0° and Iam = 17.6, 17.8, 18.0, 18.2, 18.4°, and returns the complete 5 × 5
matrix. This is a robustness analysis, not a rule for choosing positions.

```{r sens-one}
sens <- segal_sensitivity(d$two_theta, d$intensity, sample = "0%/1")
sens
```

```{r sens-all}
sens_all <- lapply(samples, function(s) {
  dd <- xrd[xrd$sample == s, ]
  segal_sensitivity(dd$two_theta, dd$intensity, sample = s)
})
names(sens_all) <- samples
ss <- do.call(rbind, lapply(sens_all, function(z) cbind(sample = z$sample, z$summary)))
knitr::kable(data.frame(Sample = ss$sample, Min = round(ss$ci_min, 2), Max = round(ss$ci_max, 2),
                        Range = round(ss$ci_range, 2), Central = round(ss$ci_central, 2)),
             caption = "Sensitivity of the Segal CI (%) to the I200/Iam positions (Range in percentage points).")
```

Two points follow from the numbers. First, the spread caused by the position
choice (5 to 22 percentage points) is larger than most of the differences
between neighbouring samples in the series, so absolute Segal values should
always be reported together with the positions used. Second, the position of the
(200) maximum within the 22-23° window is not constant across the series:

```{r apex}
apex <- do.call(rbind, lapply(samples, function(s) {
  dd <- xrd[xrd$sample == s & xrd$two_theta >= 22 & xrd$two_theta <= 23, ]
  data.frame(Sample = s, `Maximum at (°2θ)` = round(dd$two_theta[which.max(dd$intensity)], 2),
             check.names = FALSE)
}))
knitr::kable(apex, caption = "Position of the intensity maximum between 22 and 23° (diagnostic only).")
```

The maximum moves from about 22.9° to about 22.05° towards the end of the
series, which is consistent with the growing contribution of cellulose II
(whose main reflection lies at lower angle than the cellulose I (200)
reflection), as expected for mercerization (Nam et al., 2016). Because of this,
a single fixed position pair reads a different part of the profile in each
pattern, and the ordering of the samples is not preserved in every cell of the
grid (it is at the low-angle edge I200 = 22.2°):

```{r ranks}
grid <- do.call(rbind, lapply(sens_all, as.data.frame))
wide <- reshape(grid[, c("sample", "i200_position", "iam_position", "ci")],
                idvar = c("i200_position", "iam_position"), timevar = "sample",
                direction = "wide")
rho <- apply(wide[, paste0("ci.", samples)], 1,
             function(v) cor(v, all_res$ci[match(samples, all_res$sample)], method = "spearman"))
first_last <- wide[["ci.100%/1"]] < wide[["ci.0%/1"]]
data.frame(cells = 25, min_spearman_vs_central = round(min(rho), 2),
           cells_with_rho_at_least_0.85 = sum(rho >= 0.85),
           `100%/1 below 0%/1` = sum(first_last), check.names = FALSE)
```

## 9. Limitations

The Segal Crystallinity Index is an **empirical relative XRD index**. Its use
here and in general is limited by the following points (details and sources in
`?textileCrystR`):

* **Peak overlap.** Intensity at the amorphous position also contains the tails
  of broad crystalline peaks; even perfect small cellulose crystals give an index
  well below 100 %, and real samples never give 0 %.
* **Crystallite size / FWHM.** Broader peaks change the index at unchanged
  amorphous content.
* **Cellulose polymorph.** Cellulose I and II behave differently, and the
  cellulose II amorphous position underestimates the amorphous fraction.
* **Preferred orientation.** Sample form and geometry change the I200 : Iam ratio;
  reflections near 20.3-20.6° close to Iam are affected.
* **Background handling.** No background correction is applied by the package;
  values with different treatment are not comparable.
* **Choice of I200/Iam positions.** See Section 8; report the positions with the value.
* **Sample preparation.** Pressed powder, nonwoven and fabric samples, and
  spinning or geometry, can shift the index substantially.
* **Partially mercerized cotton.** For blends of cellulose I and II the (1-10) and
  (110) reflections of cellulose I remain near the cellulose II amorphous position,
  so the Segal method is not appropriate for samples that are not completely
  mercerized (Nam et al., 2016). The 22.6°/18.0° convention used here is applied
  identically to all seven patterns for comparability; for the blended samples it
  is a peak-height ratio at fixed positions, not the crystalline fraction of either
  polymorph.

Salem et al. (2023) identify four XRD approaches: Segal peak height, peak fitting
(deconvolution), amorphous subtraction and Rietveld. `textileCrystR` implements
the first two (see `vignette("getting-started")` for peak area / deconvolution);
amorphous subtraction and Rietveld integration are intentionally out of scope.

## References

Nam, S., French, A. D., Condon, B. D., Concha, M. (2016). Segal crystallinity index
revisited by the simulation of X-ray diffraction patterns of cotton cellulose Iβ and
cellulose II. *Carbohydrate Polymers* 135, 1-9.

Salem, K. S., Kasera, N. K., Rahman, M. A., Jameel, H., Habibi, Y., Eichhorn, S. J.,
French, A. D., Pal, L., Lucia, L. A. (2023). Comparison and assessment of methods
for cellulose crystallinity determination. *Chemical Society Reviews*.
doi:10.1039/d2cs00569g

Segal, L., Creely, J. J., Martin, A. E., Conrad, C. M. (1959). An empirical method
for estimating the degree of crystallinity of native cellulose using the X-ray
diffractometer. *Textile Research Journal* 29, 786-794.
